Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-02-04 Manuscript/Manuscript.tex
Add analysis model popularity data

The variability due to clustering is often measured by means of the
intra-class coefficient (ICC). The ICC can be seen as the percentage of
variance that can be attributed to the cluster-level, where a high ICC
would indicate that a lot of variability is due to the cluster
structure. Multilevel models typically accommodate for this variability
by including separate intercepts for each cluster. Such fixed effects
relieve the restriction imposed by single-level models: equal group
means across clusters. Additionally, models may include random effects
of predictors across clusters, and random error terms (heterogeneous
residual error variances; see e.g. \citet{hox17} and \citet{jong21}).
There are many names for models that take clustering into account. Some
popular examples are `multilevel models', `hierarchical models', `mixed
effect models' and `random effect models'.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations exclusively (although the alternative,
joint modeling imputation for multilevel data or \pkg{jomo}
\citet{jomo}, has been implemented in \pkg{mice} as well). Other useful
resources for the analysis of incomplete multilevel data include the
\proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and
the empirical work by \citet{audi18} and \citet{grun18}. Note that this
tutorial paper assumes a basic level of knowledge on multilevel models.
We're providing an overview of implementations. It's up-to the reader to
decide which multilevel strategy suits their data. We won't go into
detail for the different methods (and equations). This paper is just a
software tutorial, so we'll keep it practical.

@@

\begin{itemize}
\tightlist
\item
  \texttt{mice::popmis} (simulated data on perceived popularity,
  \(n = 199\) pupils across \(N = 10\) classes, with univariate MAR
  missingness);
\item
  \texttt{GJRM::hiv} (simulated data on HIV diagnoses, \(n = 6,416\)
  patients across \(N = 9\) regions, with univariate MNAR missingness)
\item
  \texttt{metamisc::impact} (empirical data on traumatic brain injuries,
  \(n = 11,022\) patients across \(N = 15\) studies, without
  \texttt{NA}s).
\end{itemize}

@@

In this section we'll go over the different steps involved with imputing
incomplete multilevel data. The data we're using is the \texttt{popmis}
dataset from the \texttt{mice} package. This is a simulated dataset with
pupils clustered in classes, where the number of units \(n = 2000\), and
the number of clusters \(N = 100\), on 7 variables:

\begin{itemize}
\tightlist
\item
  \texttt{pupil} Pupil number within class,
\item
  \texttt{class} Class number,
\item
  \texttt{extrav} Pupil extraversion,
\item
  \texttt{sex} Pupil gender,
\item
  \texttt{texp} Teacher experience (years),
\item
  \texttt{popular} Pupil popularity,
\item
  \texttt{teachpop} Teacher popularity.
\end{itemize}

The analysis model corresponding to this dataset is multilevel
regression with \texttt{popular} as the outcome variable, \texttt{sex}
and \texttt{extrav} as unit-level predictors, and \texttt{texp} and
\texttt{teachpop} as cluster-level predictors. Since the data is
simulated and the missingness is induced, we can compare our imputations
to the true data generating mechanism.

The data is created in such a way that the clustering variable
\texttt{class} explains quite some variance in the outcome variable
\texttt{popular}. We express this using the intra-class correlation, ICC
\(=\) 0.36. This tells us that it would be wise to use multilevel
modeling.

@@

@@

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the missingness depends on the
observed values of other variables. We'll highlight only one other
variable to illustrate, but ideally one would inspect all relations. The
questions we'll ask are: `Does the missing data of pupil popularity
(\texttt{popular}) depend on observed teacher popularity
(\texttt{teachpop})?'. This can be evaluated statistically, but visual
inspection usually suffices. We'll make a histogram of \texttt{teachpop}
separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that \textbf{{[}update this part{]}}
the distribution for the missing \texttt{popular} is further to the
right than the distribution for observed \texttt{popular}. This would
indicate a right-tailed MAR missingness. (In fact, this is exactly what
happens, because the missingness in these data was created manually.)
We've made it observable by examining the relations between the
missingness in popular and the observed data in \texttt{teachpop}.

@@

We'll use predictive mean matching to impute the continuous variables
and logistic regression to impute the binary variable \texttt{sex}. We
do not use the observation identifier \texttt{pupil} or cluster
identifier \texttt{class} as predictors to impute other variables.

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(popmis, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored
1  popular 0.5583543 0.4739527
2 teachpop 0.6823380 0.6823380
3     texp 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

As the original ICCs show, 100\% of the variance in \texttt{texp} can be
attributed to the clustering variable \texttt{class}. This tells us that
the multilevel structure of the data should be taken into account. If we
don't, we'll end up with incorrect imputations, biasing the effect of
the clusters towards zero.

We can also observe that the teacher experience increases slightly after
imputation. This is due to the MNAR missingness in \texttt{texp}. Higher
values for \texttt{texp} have a larger probability to be missing. This
may not a problem, however, if at least one pupil in each class has
teacher experience recorded, we can deductively impute the correct
(i.e.~true) value for every pupil in the class.

@@

We'll now use \texttt{class} as a predictor to impute all other
variables. This is still not recommended practice, since it only works
under certain circumstances and results may be biased
\citep{drec15, ende16}. But at least, it includes some multilevel
aspect. This method is also called `fixed cluster imputation', and uses
N-1 indicator variables representing allocation of N clusters as a fixed
factor in the model \citep{reit06, ende2016}. Colloquially, this is
`multilevel imputation for dummies'.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> 
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor
1  popular 0.5583543 0.4739527 0.5660108
2 teachpop 0.6823380 0.6823380 0.6823380
3     texp 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "", "", "", "", "2l.norm", "")
R> imp_norm_2l <-
+   mice(
+     popmis %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 1
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 5088 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 5088 rows containing missing values (geom_point).
\end{CodeOutput}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor      norm
1  popular 0.5583543 0.4739527 0.5660108 0.5583543
2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380
3     texp 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
R> meth <- c("", "", "", "", "", "2l.pan", "")
R> imp_pan_2l <-
+   mice(
+     popmis %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 1
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 5088 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 5088 rows containing missing values (geom_point).
\end{CodeOutput}

@@

\begin{CodeOutput}
      vars       CCA   ignored predictor      norm       pan
1  popular 0.5583543 0.4739527 0.5660108 0.5583543 0.5583543
2 teachpop 0.6823380 0.6823380 0.6823380 0.6823380 0.6823380
3     texp 1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

>>>>

The variability due to clustering is often measured by means of the
intraclass coefficient (ICC). The ICC can be seen as the percentage of
variance that can be attributed to the cluster-level, where a high ICC
would indicate that a lot of variability is due to the cluster
structure. Multilevel models typically accommodate for this variability
by including separate intercepts for each cluster. Such fixed effects
relieve the restriction imposed by single-level models: equal group
means across clusters. Additionally, models may include random effects
of predictors across clusters, and random error terms (heterogeneous
residual error variances; see e.g. \citet{hox17} and \citet{jong21}).
There are many names for models that take clustering into account. Some
popular examples are `multilevel models', `hierarchical models', `mixed
effect models' and `random effect models'.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations exclusively (although the alternative,
joint modeling imputation for multilevel data or \pkg{jomo}
\citet{jomo}, has been implemented in \pkg{mice} as well). Other useful
resources for the analysis of incomplete multilevel data include the
\proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and
the empirical work by \citet{audi18} and \citet{grun18}. Note that this
tutorial paper assumes a basic level of knowledge on multilevel models.
We're providing an overview of implementations. It's up-to the reader to
decide which multilevel strategy suits their data. We won't go into
detail for the different methods (and equations). This paper is just a
software tutorial, so we'll keep it practical. Assumed knowledge also
includes the use of the `piping operator', \texttt{\%\textgreater{}\%},
adopted from the \texttt{magrittr} package and avidly used among
\texttt{tidy} workflows, and the \texttt{lme4} notation for multilevel
models.

@@

\begin{itemize}
\tightlist
\item
  \texttt{mice::popmis} (simulated data on perceived popularity,
  \(n = 2,000\) pupils across \(N = 100\) schools, with univariate MAR
  missingness);
\item
  \texttt{GJRM::hiv} (simulated data on HIV diagnoses, \(n = 6,416\)
  patients across \(N = 9\) regions, with univariate MNAR missingness)
\item
  \texttt{metamisc::impact} (empirical data on traumatic brain injuries,
  \(n = 11,022\) patients across \(N = 15\) studies, without
  \texttt{NA}s).
\end{itemize}

@@

In this section we'll go over the different steps involved with imputing
incomplete multilevel data. The data we're using is the \texttt{popmis}
dataset from the \texttt{mice} package. This is a simulated dataset with
pupils (\(n = 2000\)) clustered within schools (\(N = 100\)). In this
tutorial we'll use the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{school}, school identification number (clustering variable);
\item
  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
  unit-level);
\item
  \texttt{sex}, pupil sex (0=boy, 1=girl; unit-level);
\item
  \texttt{texp}, teacher experience (in years; cluster-level).
\end{itemize}

The analysis model corresponding to this dataset is multilevel
regression with random intercepts, random slopes and a cross-level
interaction. The outcome variable is \texttt{popular}, which is
predicted from the unit-level variable \texttt{sex} and the
cluster-level variable \texttt{texp}. The regression equation and
\texttt{lme4} notation for this model are

\[
\text{popular}_{ij} =
\gamma_{00} + 
\gamma_{10} \text{ sex}_{ij} + 
\gamma_{01} \text{ texp}_{j} + 
\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
u_{0j} + 
u_{1j} \text{ sex}_{ij} + 
e_{ij} \\
\]

\[
\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
\]

Since the data is simulated and the missingness is induced, we can
compare our inferences after imputation to the true complete data. The
data is created in such a way that the clustering variable
\texttt{school} explains quite some variance in the outcome variable
\texttt{popular}. We express this using the intraclass correlation, ICC
\(=\) 0.58. We'll evaluate the ICC after each missing data strategy, and
compare the estimated fixed effect of \texttt{sex} on \texttt{popular}
\(\gamma_{10} = 0.843\).

@@

Load the data into the environment and select the relevant variables:

\begin{CodeChunk}
\begin{CodeInput}
R> popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
\end{CodeInput}
\end{CodeChunk}

@@

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the missingness depends on the
observed values of other variables. We'll highlight only one other
variable to illustrate, but ideally one would inspect all relations. The
questions we'll ask are: `Does the missing data of pupil popularity
(\texttt{popular}) depend on observed teacher popularity
(\texttt{texp})?'. This can be evaluated statistically, but visual
inspection usually suffices. We'll make a histogram of \texttt{texp}
separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that \textbf{{[}update this part{]}}
the distribution for the missing \texttt{popular} is further to the
right than the distribution for observed \texttt{popular}. This would
indicate a right-tailed MAR missingness. (In fact, this is exactly what
happens, because the missingness in these data was created manually.)
We've made it observable by examining the relations between the
missingness in popular and the observed data in \texttt{texp}.

@@

We'll use predictive mean matching to impute the continuous variables
and logistic regression to impute the binary variable \texttt{sex}. We
do not use the observation identifier \texttt{pupil} or cluster
identifier \texttt{school} as predictors to impute other variables.

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(popmis, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("school")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

As the original ICCs show, 100\% of the variance in \texttt{texp} can be
attributed to the clustering variable \texttt{school}. This tells us
that the multilevel structure of the data should be taken into account.
If we don't, we'll end up with incorrect imputations, biasing the effect
of the clusters towards zero.

We can also observe that the teacher experience increases slightly after
imputation. This is due to the MNAR missingness in \texttt{texp}. Higher
values for \texttt{texp} have a larger probability to be missing. This
may not a problem, however, if at least one pupil in each school has
teacher experience recorded, we can deductively impute the correct
(i.e.~true) value for every pupil in the school.

@@

We'll now use \texttt{school} as a predictor to impute all other
variables. This is still not recommended practice, since it only works
under certain circumstances and results may be biased
\citep{drec15, ende16}. But at least, it includes some multilevel
aspect. This method is also called `fixed cluster imputation', and uses
N-1 indicator variables representing allocation of N clusters as a fixed
factor in the model \citep{reit06, ende2016}. Colloquially, this is
`multilevel imputation for dummies'.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> # pred[, "pupil"] <- 0
R> 
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(-2, 2, 2, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "2l.norm", "",  "")
R> imp_norm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 848 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 848 rows containing missing values (geom_point).
\end{CodeOutput}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(-2, 2, 1, 2)
R> meth <- c("", "2l.pan", "", "")
R> imp_pan_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Warning: Removed 848 rows containing non-finite values (stat_boxplot).
\end{CodeOutput}
\begin{CodeOutput}
Warning: Removed 848 rows containing missing values (geom_point).
\end{CodeOutput}

@@
